Visualizing Contingency Tables

### 2006-11-17 trunk/vcd/paper/Shadings/JCGS.Rnw
finished draft version for colors paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@590 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
fully integrated new hcl2hex into papers
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@620 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}
@

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS color space.
The implementation in the standard packages of the
\proglang{R} system for statistical computing and graphics
\citep{vcd:R:2005} employs colors from HSV space. 
Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. As both spaces can easily generate the
same colors, we only discuss HSV space in the following.

@@

## full colors
rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))

text(-100, 33, "full color", pos = 4, cex = 1.2)
rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
rect(0, 20, 20, 30, col = hcl(260, 0, 90))
#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
rect(40, 20, 60, 30, col = hcl(260, 50, 70))
rect(80, 20, 100, 30, col = hcl(260, 100, 50))

@@

## reduced colors
rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))

text(-100, 13, "reduced color", pos = 4, cex = 1.2)
rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
rect(0, 0, 20, 10, col = hcl(260, 0, 90))
rect(20, 0, 40, 10, col = hcl(260, 10, 70))
rect(40, 0, 60, 10, col = hcl(260, 20, 50))

>>>>

hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl2hex(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}
@

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS color space.
The implementation in the standard packages of the
\proglang{R} system for statistical computing and graphics
\citep{vcd:R:2006} employs colors from HSV space. 
Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. As both spaces can easily generate the
same colors, we only discuss HSV space in the following.

@@

## full colors
rect(-100, 47.5, -90, 52.5, col = hcl2hex(0, 100, 50))
rect(-55, 67.5, -45, 72.5, col = hcl2hex(0, 50, 70))
rect(-5, 95, 5, 100, col = hcl2hex(260, 0, 100))       ## grey vs. white
rect(-5, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))     ## grey vs. white
rect(45, 67.5, 55, 72.5, col = hcl2hex(260, 50, 70))
rect(90, 47.5, 100, 52.5, col = hcl2hex(260, 100, 50))

text(-100, 33, "full color", pos = 4, cex = 1.2)
rect(-100, 20, -80, 30, col = hcl2hex(0, 100, 50))
rect(-60, 20, -40, 30, col = hcl2hex(0, 50, 70))
rect(-20, 20, 0, 30, col = hcl2hex(0, 0, 90))       
rect(0, 20, 20, 30, col = hcl2hex(260, 0, 90))
#white# rect(-20, 20, 0, 30, col = hcl2hex(0, 0, 100))
#white# rect(0, 20, 20, 30, col = hcl2hex(260, 0, 100))
rect(40, 20, 60, 30, col = hcl2hex(260, 50, 70))
rect(80, 20, 100, 30, col = hcl2hex(260, 100, 50))

@@

## reduced colors
rect(-25, 47.5, -15, 52.5, col = hcl2hex(0, 20, 50))
rect(-15, 67.5, -5, 72.5, col = hcl2hex(0, 10, 70))
rect(5, 67.5, 15, 72.5, col = hcl2hex(260, 10, 70))
rect(25, 47.5, 15, 52.5, col = hcl2hex(260, 20, 50))

text(-100, 13, "reduced color", pos = 4, cex = 1.2)
rect(-60, 0, -40, 10, col = hcl2hex(0, 20, 50))
rect(-40, 0, -20, 10, col = hcl2hex(0, 10, 70))
rect(-20, 0, 0, 10, col = hcl2hex(0, 0, 90))
rect(0, 0, 20, 10, col = hcl2hex(260, 0, 90))
rect(20, 0, 40, 10, col = hcl2hex(260, 10, 70))
rect(40, 0, 60, 10, col = hcl2hex(260, 20, 50))
